Quantitative Evaluation of Touch-Screen Performance with Transition Matrix Measures

### 2015-08-31 flux-entropy-paper/flux-entropy-paper.tex
lots of new content for the paper

@@

In the following sections, we will use transition matrices, flux, and
entropy to analyse a corpus of data collected as part of our research
into collaborative interactions with musical touch-screen tablet
interfaces. In a period of more than two years (April 2013 - June 2015), we have
collected data from $139$ sessions where various participants used
musical touch-screen interfaces on Apple iPad devices. Each session
took place in a number of contexts including rehearsals, performances, lab study
sessions, classes, and casual demonstrations. Participants performed
free improvisations as well as written compositions and performed on
iPads alone or using setups of iPads and acoustic percussion
instruments. The median length of these sessions was 7m32s (minimum was
1m0s and maximum was 26m) and the
median number of participants was four (minimum 1, maximum 9).

\section{Modelling Performer Ratings with Flux and Entropy}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/stat-response-fit.pdf}
  \caption{Transition matrix statistics (flux above and entropy below) vs Likert responses for
    questions 3, 6 and 7, all of which exhibited significant
    relationships under a proportional odds logistic regression model.
    A least-squares linear model is shown with a dotted line.
    \label{fig:stat-response-fit}}
\end{figure}

One final question to explore is how these flux and entropy measures
in the interaction data relate to the performers' perceptions of their
performance?

To answer this question, we construct models with the flux and entropy
measures as independent variables and the Likert responses (1 to 5) as
the dependent variable. These measures exhibited clear relationships
for three of the survey questions (Questions 3, 6 and 7), which are
plotted in Figure~\ref{fig:stat-response-fit} where the least-squares
line of best fit has a positive gradient for both flux and entropy.

Determining the strengths of these positive relationships is
difficult, both due to the problem of treating ordinal Likert
responses as numeric~\cite{Gardner:2007fj} and also
due to the unknown distribution of the entropy and flux statistics. As
a result, we use a conservative proportional odds logistic regression
(POLR)~(using the \texttt{polr} function in \texttt{R}, from the MASS
package \cite{Venables:2002qv}) to treat the Likert responses as an
ordered factor response.

The POLR results indicate a significant ($p<0.05$) relationship
between the user-reported Likert response and the flux statistic for
Question 3 ($t = 2.77, df = 4, p = 0.025$), Question 6
($t = 2.41, df = 4, p = 0.037$), and Question 7
($t = 2.49, df = 4, p = 0.034$). The same questions displayed a
significant relationship for the entropy statistic: Question 3
($t = 2.93, df = 4, p = 0.022$), Question 6
($t = 2.69, df = 4, p = 0.028$), and Question 7
($t = 2.89, df = 4, p = 0.022$). These relationships are shown in
Figure~\ref{fig:stat-response-fit}.

It is notable that these questions (3, 6, and 7) all pertain to the
\emph{perceived effect of the agent} on the performers. As the agent
uses the flux measure to trigger new-idea messages, it appears that
performers appreciated its intervention more in particularly creative
performances where flux and entropy were high.

>>>>

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figures/three-performance-contexts}
  \caption{Collaborative musical interaction with our touch-screen
    interfaces in three different contexts: (left to right) rehearsal
    in a studio,
    performance in front of an audience, and demonstration at a
    research poster session. All of these session were improvised,
    and in the performance session, the musicians used percussion
    instruments as well as iPads.
    \label{fig:three-performance-contexts}}
\end{figure*}

@@

\begin{table}
%\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 150       &         &        &  \\
Record size & 306MB     &         &        &   \\
Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
Participants& 524       & 1       & 4      & 9    \\
Flux        &           & 0.02239 & 0.1445 & 0.3246\\
\hline
\end{tabular}
\caption{
  Representative data about the corpus of musical touch-screen
  interaction sessions used in this study. These sessions were
  rehearsals, performances, studies, and demonstration sessions and the
  participants improvised or performed written compositions. Each record
  consisted of touch-data recorded in CSV format and was later analysed
  with a gestural classification system. Some participants were
  present in multiple sessions, however this data was not tracked in
  the corpus.\label{corpus-table}}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/sessions-count}
  \caption{Distribution of the sessions in our interaction corpus by
    performance context. Improvisations or compositions were repeated
    several times in rehearsals and study sessions but only performed
    once in concerts.
    \label{fig:count-data}}
\end{figure}

In the following sections, we will use transition matrices and the
flux measure to analyse a corpus of data collected as part of our research
into collaborative interactions with musical touch-screen tablet
interfaces.

Over a period of more than two years (April 2013 - June 2015), we have
collected data from $150$ sessions where various participants used
musical touch-screen interfaces on Apple iPad devices. These
interfaces all shared a common mode of interaction where the majority
of the screen could be used for free form touch interactions; tapping
produced short sounds, while swiping or swirling produced long sounds
with volume proportional to the velocity of the moving touch point.
All touch interactions from each participant in a session was recorded
by sending data to a central server application. This data was stored
in CSV files recording the date and time, X, and Y locations,
velocity, and a unique device identifier for each touch event detected
by the iOS operating system.

Each session took place in a number of contexts including rehearsals,
performances, lab study sessions, classes, and casual demonstrations,
some of which are shown in Figure
\ref{fig:three-performance-contexts}. Participants performed free
improvisations as well as written compositions and performed on iPads
alone or using setups of iPads and acoustic percussion instruments.
The median length of these sessions was 7m29s and the median number of
participants was four, other descriptive statistics about the sessions
is shown in Table \ref{corpus-table}. Each session was coded with
metadata regarding the session context (whether it was a performance,
rehearsal, study, demonstration, or data collection session), the
session type (either improvisation or composition), and the
instruments used (iPads alone, or iPads with acoustic percussion). The
distribution of recorded sessions by context and type is shown in
Figure \ref{fig:count-data}.

To analyse the gestural interactions present in each of these
sessions, the sessions were automatically coded according to a
vocabulary of nine simple, continuous, touch-screen gestures. Examples
of these gestures was generated by an expert user in formal process
where each gesture was performed for 1-minute and feature vectors of
discriptive statistics on these examples was used to train a Random
Forest~\cite{Breiman:2001kx} classifier\footnote{The classifier used
  was from Python's Scikit-learn package~\cite{scikit-learn}.}. The
classifier was applied to 5-second windows of the touch-data each
second provide a chain of gesture-states for each participant in each
session. These chains can be used to create $9 /times 9$
gesture-transition matrices for individual performers, and the whole
ensemble in each session.

% While musical interactions such as those in
% our corpus are often analysed using qualitative methods, it is
% problematic to

While the interaction sessions recorded in our corpus were performed
by many different participants with varying experience levels, in
different contexts and on different instrumental setups, the common
interface and method of recording the data allows all the sessions to
be transformed into gesture sequences. This common format allows us to
ask if performance styles vary in between these different situations.
The sequences can be plotted and inspected qualitatively, but
transition matrices present a quantitative approach to analysing such
interactions.

In the rest of this paper we will use our matrix measure techniques to
interrogate this dataset and to answer two questions. First, do the
transition matrix measures allow us to distinguish between different
performance contexts, types and instrumental setups? Secondly, do
these performance measure allow us further insight into the corpus
than the metadata can provide?

\subsection{Differentiating interaction sessions}
\label{differentiating-interaction-sessions}

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figures \ref{fig:flux-boxplot},
\ref{fig:entropy-boxplot}, and \ref{fig:flux-entropy-distribution}. A
three-way ANOVA procedure was performed on the data set to find
whether the metatdata could significantly predict the entropy and flux
values. In the following analysis, only the 95 performative sessions
were considered (i.e. not the demonstration or data collection
contexts).

All main effects on flux were found to be significant: performance
context ($F(2,86) = 6.28, p < 0.01$), performance type ($F(1,86) =
7.21, p < 0.01$), and instrumentation ($F(1,86) = 20.06, p < 0.001$).
Significant interaction effects were also found for performance
context and type ($F(1,86) = 6.18, p < 0.05$), and performance context and
instrumentation ($F(1,86) = 10.48, p < 0.01$).

For entropy, significant main effects were found due to performance
context ($F(2,86) = 12.10, p < 0.001$), and instrumentation ($F(1,86)
= 25.86$). The interaction of performance type and instrumentation was
also found to be significant ($F(1,86) = 9.93, p<0.01$).

These tests show that the different interaction styles that might be
present in performance contexts, types and
instrumentations are reflected in the values of flux and entropy of
their transition matrices.

% need to add overall contrasts column to do pairwise t-tests.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/flux-boxplot}
  \caption{Distributions of flux values by performance context. Study
    and rehearsal sessions had the highest flux for improvisations
    which fell during performances. Compositions had lower flux in
    rehearsals than performances.
    \label{fig:flux-boxplot}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/entropy-boxplot}
  \caption{Distribution of entropy by
  performance context. Demonstration sessions had much lower entropy
  indicating data unevenly spread in the transition matrices. This
  could be because untrained
  participants did not explore the full gestural potential of the
  instruments. 
  \label{fig:entropy-boxplot} }
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/flux-entropy-distribution}
  \caption{Distribution of flux and entropy by performance type and
    context. Flux and entropy tended to be linearly related in
    improvisations, particularly in demonstrations and studies, while
    compositions had a more scattered relationship.
    \label{fig:flux-entropy-distribution}}
\end{figure}

\subsection{Breaking performances into sections}

It is clear that flux and entropy can detect the variation in
performance styles in between performances of different types, but can
they be used to understand some of the internal structure of
performances? A simple model for understanding temporal art-forms is
the ``beginning, middle, end'' structure. To investigate how the
structure fits session in our corpus, we can divide the
gestural-sequences of each session into thirds and calculate the
transition matrices given by each section.

\subsection{Performance styles over time}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/flux-through-time}
  \caption{Distribution flux values of sessions through time.
    Different performance styles are visible as clusters of similar
    flux regardless of performance context.
    \label{fig:flux-through-time}}
\end{figure}

\section{Conclusions}

% \section{Modelling Performer Ratings with Flux and Entropy}

% \begin{figure}
%   \centering
%   \includegraphics[width=\linewidth]{figures/stat-response-fit.pdf}
%   \caption{Transition matrix statistics (flux above and entropy below) vs Likert responses for
%     questions 3, 6 and 7, all of which exhibited significant
%     relationships under a proportional odds logistic regression model.
%     A least-squares linear model is shown with a dotted line.
%     \label{fig:stat-response-fit}}
% \end{figure}

% One final question to explore is how these flux and entropy measures
% in the interaction data relate to the performers' perceptions of their
% performance?

% To answer this question, we construct models with the flux and entropy
% measures as independent variables and the Likert responses (1 to 5) as
% the dependent variable. These measures exhibited clear relationships
% for three of the survey questions (Questions 3, 6 and 7), which are
% plotted in Figure~\ref{fig:stat-response-fit} where the least-squares
% line of best fit has a positive gradient for both flux and entropy.

% Determining the strengths of these positive relationships is
% difficult, both due to the problem of treating ordinal Likert
% responses as numeric~\cite{Gardner:2007fj} and also
% due to the unknown distribution of the entropy and flux statistics. As
% a result, we use a conservative proportional odds logistic regression
% (POLR)~(using the \texttt{polr} function in \texttt{R}, from the MASS
% package \cite{Venables:2002qv}) to treat the Likert responses as an
% ordered factor response.

% The POLR results indicate a significant ($p<0.05$) relationship
% between the user-reported Likert response and the flux statistic for
% Question 3 ($t = 2.77, df = 4, p = 0.025$), Question 6
% ($t = 2.41, df = 4, p = 0.037$), and Question 7
% ($t = 2.49, df = 4, p = 0.034$). The same questions displayed a
% significant relationship for the entropy statistic: Question 3
% ($t = 2.93, df = 4, p = 0.022$), Question 6
% ($t = 2.69, df = 4, p = 0.028$), and Question 7
% ($t = 2.89, df = 4, p = 0.022$). These relationships are shown in
% Figure~\ref{fig:stat-response-fit}.

% It is notable that these questions (3, 6, and 7) all pertain to the
% \emph{perceived effect of the agent} on the performers. As the agent
% uses the flux measure to trigger new-idea messages, it appears that
% performers appreciated its intervention more in particularly creative
% performances where flux and entropy were high.
